Fast gravitational wave parameter estimation without compromises

### 2023-01-26 src/tex/ms.tex
Add spline flow sentence and maximum likelihood with search parameters

\begin{algorithm}
\caption{\textsc{flowMC} pseudocode}\label{alg:cap}
\KwIn{initial position $ip$}
\Parameters{number of training loops $nt$, number of production loops $np$}
\Variables{current chain $cc$, current position $cp$, current NF parameters $\Theta$}
\KwResult{$chains$}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$\\
        $chains, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $cc \leftarrow Append(cc, chains)$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $c_{local}, cp \leftarrow LocalSampling(cp)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $chains \leftarrow Append(chains, c_{local}, c_{global})$
    }

@@

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 seconds of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz.  We draw noise from the design
PSDs for the LIGO Hanford, LIGO Livingston
(\textsc{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\textsc{SimNoisePSDAdvVirgo}) detectors \cite{}.
For both injection and recovery, we make use of the \textsc{IMRPhenomD} waveform
\cite{Khan:2015jqa} via the fully-differentiable implementation presented in the
\textsc{ripple} package \cite{}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
for the GW150914 analysis, and 128 s of data sampled at 4096 Hz for the
GW170817 analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. For our specific choice of sampler settings, we produce
${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using
\textsc{arviz} \cite{arviz_2019},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 10 minutes. Most of this
time is spent on just-in-time (JIT) compilation of the code; the actual
sampling time is only ${\sim}150$ s. We pre-compute the reference waveform
parameters used in hetetrodyne likelihood for the two events, the time for
which is omitted in the \texttt{wall} time calculation.  The chain data and the
analysis scripts which generate the chains can be found in 
% \kw{\variable{data/} and \variable{script}}.

@@

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\textsc{differential evolution} algorithm we currently use has not been
implemented in \textsc{Jax}, and the \textsc{Jax} waveform we use is not
compatible with the parallelization scheme in the \textsc{scipy} library,
maximizing the likelihood currently takes us around 5 minutes \kw{Benchmark it
in the code and make this number precise.} for GW170817. There are two ways to
reduce this time.

@@

Second, we may marginalize extrinsic parameters to reduce the dimensionality of
the optimization problem. Currently, we simultaneously maximize all 11 CBC
parameters in our problem numerically, which is unnecessary. There are
long-existing, efficient maximization schemes for extrinsic parameters, such as
the merger time and phase, which can find the corresponding maximum likelihood
waveform much more efficiently when compared to differential evolution. We
expect implementing these schemes will reduce the time needed
to find the reference waveform parameters by fixing the extrinsic parameters
and by reducing the dimensionality of the optimization problem.

>>>>

\begin{algorithm}
\caption{\textsc{flowMC} pseudocode}\label{alg:cap}
\KwIn{initial position $ip$}
\Parameters{number of training loops $nt$, number of production loops $np$}
\Variables{current chains $cc$, current position $cp$, current NF parameters $\Theta$, chains from local sampler $c_{local}$, chains from global sampler $c_{global}$}
\KwResult{output chains $chains$}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $cc \leftarrow Append(cc, c_{global})$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $c_{local}, cp \leftarrow LocalSampling(cp)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $chains \leftarrow Append(chains, c_{local}, c_{global})$
    }

@@

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 seconds of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz.  We draw noise from the design
PSDs for the LIGO Hanford, LIGO Livingston
(\textsc{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\textsc{SimNoisePSDAdvVirgo}) detectors \cite{}. For both injection and
recovery, we make use of the \textsc{IMRPhenomD} waveform \cite{Khan:2015jqa}
via the fully-differentiable implementation presented in the \textsc{ripple}
package \cite{}. We use a neural spline flow model \cite{2019arXiv190604032D}
with 10 layers, each with 128 hidden units, and 8 bins per layer as our
normalizing flow model.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply it
to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors shown in
Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz for the
GW150914 analysis, and 128 s of data sampled at 4096 Hz for the GW170817
analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. We use the same normalizing flow model as the injection-recovery
study. For our specific choice of sampler settings, we produce ${\sim}$2500 and
3500 \emph{effective samples}\footnote{Effective samples here refers to the
number of independent samples, which is the total number of generated samples
divided by their correlation length; we compute the effective sample size using
\textsc{arviz}
\cite{arviz_2019},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 10 minutes. Most of this time is
spent on just-in-time (JIT) compilation of the code; the actual sampling time is
only ${\sim}150$ s. We pre-compute the reference waveform parameters used in
hetetrodyne likelihood for the two events, the time for which is omitted in the
\texttt{wall} time calculation.

@@

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\textsc{differential evolution} algorithm we currently use has not been
implemented in \textsc{Jax}, and the \textsc{Jax} waveform we use is not
compatible with the parallelization scheme in the \textsc{scipy} library,
maximizing the likelihood currently takes us around 1 minute for GW170817. There
are two ways to reduce this time.

@@

Second, we may marginalize extrinsic parameters to reduce the dimensionality of
the optimization problem. Currently, we simultaneously maximize all 11 CBC
parameters in our problem numerically, which is unnecessary. There are
long-existing, efficient maximization schemes for extrinsic parameters, such as
the merger time and phase, which can find the corresponding maximum likelihood
waveform much more efficiently when compared to differential evolution. We
expect implementing these schemes will reduce the time needed to find the
reference waveform parameters by fixing the extrinsic parameters and by reducing
the dimensionality of the optimization problem. Furthermore, the search pipeline
provides a subset of number such as the masses, which can be fixed during the
optimization to further reduce the dimensionality of the problem.
